---
title: "Prediction Assignment Writeup"
author: "Abinashi Prakash"
date: "28 May 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

###Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants as they perform barbell lifts correctly and incorrectly 5 different ways.

Six young healthy participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:
* Class A - exactly according to the specification
* Class B - throwing the elbows to the front
* Class C - lifting the dumbbell only halfway
* Class D - lowering the dumbbell only halfway
* Class E - throwing the hips to the front


###Data

The training data for this project are available at:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available at:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

###GOAL

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

###Function and Data Download

```{r Load_Var,echo=TRUE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(e1071)
library(randomForest)
set.seed(1)

train.url <-
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- 
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

path <- paste(getwd(), sep="")
train.file <- file.path(path, "machine-train-data.csv")
test.file <- file.path(path, "machine-test-data.csv")

if (!file.exists(train.file)) {
        download.file(train.url, destfile=train.file)
}
if (!file.exists(test.file)) {
        download.file(test.url, destfile=test.file)
}

train.data.raw <- read.csv(train.file, na.strings=c("NA","#DIV/0!",""))
test.data.raw <- read.csv(test.file, na.strings=c("NA","#DIV/0!",""))

```

###Remove unecessary colums

```{r Remove_Columns,echo=TRUE}
# Drop the first 7 columns as they're unnecessary for predicting.
train.data <- train.data.raw[,8:length(colnames(train.data.raw))]
test.data <- test.data.raw[,8:length(colnames(test.data.raw))]

# Drop colums with NAs
train.data <- train.data[, colSums(is.na(train.data)) == 0] 
test.data <- test.data[, colSums(is.na(test.data)) == 0] 

# Check for near zero variance predictors and drop them if necessary
nzv <- nearZeroVar(train.data,saveMetrics=TRUE)
zero.var.ind <- sum(nzv$nzv)

if ((zero.var.ind>0)) {
        train.data <- train.data[,nzv$nzv==FALSE]
}
```

###Slice the data for cross validation
The training data is divided into two sets. This first is a training set with 70% of the data which is used to train the model. The second is a validation set used to assess model performance.

```{r Slice_Data, echo=TRUE}
in.training <- createDataPartition(train.data$classe, p=0.70, list=F)
train.data.final <- train.data[in.training, ]
validate.data.final <- train.data[-in.training, ]
```

##Model Development

##Train the model
The training data-set is used to fit a Random Forest model. A Random Forest algorithm is a way of averaging multiple deep decision trees, trained on different parts of the same data-set, with the goal of reducing the variance. This typically produces better performance at the expense of bias and interpret-ability. The Cross-validation technique assesses how the results of a statistical analysis will generalize to an independent data set. In 5-fold cross-validation, the original sample is randomly partitioned into 5 equal sized sub-samples. a single sample is retained for validation and the other sub-samples are used as training data. The process is repeated 5 times and the results from the folds are averaged.

```{r Model,echo=TRUE}
control.parms <- trainControl(method="cv", 5)
rf.model <- train(classe ~ ., data=train.data.final, method="rf",trControl=control.parms, ntree=251)
rf.model
```


###Estimate performance

```{r Predict,echo=TRUE}
rf.predict <- predict(rf.model, validate.data.final)
confusionMatrix(validate.data.final$classe, rf.predict)

accuracy <- postResample(rf.predict, validate.data.final$classe)
acc.out <- accuracy[1]

overall.ose <- 
        1 - as.numeric(confusionMatrix(validate.data.final$classe, rf.predict)
                       $overall[1])
```

###Run the model
```{r Show_Model,echo=TRUE}
results <- predict(rf.model, 
                   test.data[, -length(names(test.data))])
results
##Decision Tree
treeModel <- rpart(classe ~ ., data=train.data.final, method="class")
fancyRpartPlot(treeModel)
```
